Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-06-29 MIMOSA.tex
Discussion and conclusion. working on SI
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}. These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}. In the context of single-cell gene expression, very few methods have been proposed. To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements. However, the proportion of single cells expressing individual genes also carries information and should be evaluated.

@@

\textit{Flow cytometry}: This dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:data} for details)~\cite{Peiperl:2010ej}. The statistical analysis of responder\footnote{need to define this} and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}. The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of flow-sorted CD8 T-cells from sixteen individuals. The T-cells from blocks of four individuals were stimulated with different antigens, gene expression measured at the single-cell level and compared to paired, unstimulated controls. \footnote{Add a bit more info here, how many genes, tetramer sorted, etc}.

@@

\subsection{Beta-Binomial Model for One Marker}
For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,

@@

\subsection{Parameter estimation}
\label{sec:estimation}
Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, to simplify our calculations. Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Supplementary material), and are given by,
 \begin{align}
  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
	&=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
 \end{align} 
and
\begin{align}
	\begin{split}
\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}.\label{model2:unconstrained}
\end{split}
\end{align}
Assuming that the missing data, the $z_i$'s, are known, we define the complete log-likelihood as follows,
\begin{equation}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w)
\label{eq:cll}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated.

@@

\subsubsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
Details about the algorithms are given in Supplementary material. We used the method of Raftery and Lewis \citep{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

@@

The MMOSA model had higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non--responders when the magnitude of response was small (i.e. ENV-1-PTEG stimulation) (Figure~\ref{fig:HVTN065} A),  (AUC=0.954 for MIMOSA vs. 0.923 for Fisher's exact test vs. 0.827 for LRT vs. 0.775 for ranked fold change). When the magnitude of response was larger (GAG-1-PTEG stimulation), MIMOSA performed as well as Fisher's exact test, and better than other competing methods (Figure~\ref{fig:HVTN065} C) (AUC=0.982 for MIMOSA vs. 0.978 for Fisher's exact test vs. 0.877 for LRT vs. 0.784 for ranked fold change). Irrespective of the magnitude of the response to stimulation, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B,D). These results are consistent for other cytokines (see supplementary material).

\subsection{Fluidigm}
\subsubsection{MIMOSA Identifies Stimulation-Specific Patterns of Gene Expression in Fluidigm Single--Cell Data}
We applied the MIMOSA model to Fluidigm single--cell gene expression data from CD8+ T--cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples. The unconstrained MIMOSA model was fit to each stimulation. From the model we calculated the posterior probabilities of response for each gene, the posterior differences in the proportion of cells expressing each gene between stimulated and unstimulated conditions, as well as the posterior log ratio of the proportion of cells expressing each gene in stimulated and unstimulated conditions. The data are presented in (Figure~\ref{fig:fluidigm} A-C). Importantly, we see that MIMOSA identifies stimulation--specific differences in the proportions of cells expressing each gene, while preserving inter--subject variability (Figure~\ref{fig:fluidigm}). These patterns are preserved in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), as well as in the posterior differences and log--odds (Figure~\ref{fig:fluidigm} B,C).

@@

\subsection{Parameter estimation}
Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (See supplementary material). For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
\begin{align}
\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
\end{align}
where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$. Similarly the marginal likelihood for the alternative model is given by 
\begin{align}
\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}.\label{eq:postmult}
\end{align}
The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger. In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than 2. The EM algorithms becomes very dependent on the initial values, and can even fail to converge when good initial values are provided. Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and can be used even when $K$ is large. More details about both algorithms are given in Supplementary material.

@@

Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we performed simulations (five--dimensional data) to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, Fisher's exact test performed marginally on each dimension of the data and p--values combined through Fisher's method ($\chi^2=-2\sum_{i=1}^k\log(p_i)$), as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C). These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data.

@@

Importantly, the analysis of real--world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}). Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination but the model was blinded to vaccine time point. Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre--vaccination) and 182 (post--vaccination) than the current standard approach (Fisher's exact test), or non--Bayesian alternatives such as the likelihood ratio test or ranking by log--fold change (Figure~\ref{fig:HVTN065} A,C). Our approach treats all day 0 observations as true negatives and all day 182 observations as true positives, yet we know that not all vaccine recipients are likely to exhibit a vaccine response on day 182. None the less, this shortcoming, at worst, leads us to underestimate the true sensitivity of MIMOSA, and does not negatively impact the comparison. Additionally, MIMOSA was found to be a better fit to real--world ICS data from vaccine trials than other analysis approaches, as evidenced by more accurate estimates of true false discovery rate (Figure~\ref{fig:HVTN065} B,D).

%We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).  For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test. These two observations had a 2.32 and 3.08 fold increase in the proportion of cytokine--positive cells in the stimulated vs unstimulated samples, respectively, but differences in absolute positive cell counts of 2-3 cells, although the negative cell counts for the simulated and unstimulated samples were substantially different (32K vs 82K cells, and 65K vs 145K cells, respectively), resulting in the significant difference in proportions. This is a clear example where sharing information through the empirical Bayes framework, and eliminating the conditioning on fixed margins inherent in Fisher's exact test, lead to improved sensitivity on real data for the MIMOSA model. This increased power to detect vaccine responders is important since it impacts decision--making about which vaccine candidates move forward to later phase clinical trials.

Measuring differences in polyfunctional cell populations (i.e. identifying changes in cell populations that co--express multiple proteins, cytokines, or genes), is important in single cell immunological assays since they allow the identification of functionally distinct cell subsets in heterogeneous populations~\cite{Milush:2009bz}. In vaccine trials, these polyfunctional cell subsets may be correlated with differential vaccine response or efficacy and have the potential to inform vaccine development and trial design. Existing methods for identifying polyfunctional cytokine profiles from ICS data either separately test individual combinations of cytokines or perform global tests between groups of individuals even when the populations under study are known to be heterogeneous (i.e. mixtures of responders and non--responders).  Sometimes tests are empirical in nature (i.e. thresholds based on fold change or some factor over background), and these methods generally do not share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}. As others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account the cytokine combinations of interest~\cite{Nason:2006dx}. A global test, one that tests for a difference between all proportions of cytokine expressing cells will have lower power than one testing for difference in a specific cytokine combination if only that specific combination is truly different. This approach is inherent in our mixture model framework, where each mixture component representing some pattern of cytokine expression that is of interest.

We show a proof of principle example with two components to identify responder and non--responder observations for the four--dimensional (two cytokine) case. Sharing of parameters across nested models in the two--cytokine case allows us to reduce the number of parameters. The model is easily extended. First, to multiple components to identify not just responders and non--responders, but responders with respect to each combination of cytokines. Second, extensions to multiple cytokines are straightforward, only requiring increasing the dimensionality of the Multinomial--Dirichlet distribution. With the increase in dimensionality there is a combinatorial explosion in the number of combinations of cytokines, and the resulting number of parameters in the model. We present several possible approaches to reducing the number of parameters in the model by sharing the hyperparamters across different combinations of cytokines. A number of these strategies can be generalized to models with multiple components that allow us to identify observations with differences in specific cytokine combinations between treatment and control samples.

\section{Conclusions}
We have developed a combinatorial mixture model framework for identifying differences between treatment conditions in paired observations of cell counts from a variety of single--cell assays. The software is implemented in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
%   \includegraphics[width=.75\columnwidth]{Figures/12} 
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.6\columnwidth]{Figures/2}};
    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.6\columnwidth]{Figures/12}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
    \node at (0,-0.25) [font=\tiny\sffamily] {C} ;
    \node at (4.6,-0.25) [font=\tiny\sffamily] {D} ;
\end{tikzpicture}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) GAG-1-PTEG and expressing IL2 and IFNg. ROC and FDR plots of other cytokine combinations can be found in the Supplementary material.}
\label{fig:HVTN065}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
\node[anchor=south west, inner sep=0] at (4.6,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
\node[anchor=south west, inner sep=0] at (9.1,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot3.pdf}};
\node at (0,3) [font=\tiny\sffamily] {A} ;
\node at (4.5,3) [font=\tiny\sffamily] {B} ;
\node at (8.95,3) [font=\tiny\sffamily] {C} ;
\end{tikzpicture}
\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a significant decrease, green a significant increase, relative to the control). Columns and rows are clustered based on these signed posterior probabilities. B) The $log_2$ ratio of the proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. C) The difference in the proportion of cells expressing each gene in the stimulated vs. control samples. Ordering of the rows and columns is preserved as in A). The traces show the deviations of each cell from zero.}
\label{fig:fluidigm}
\end{figure}

>>>>

Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that. Added ref for the HVTN054 trial and the ICS standardization paper from Horton})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej}. These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}. In the context of single-cell gene expression, very few methods have been proposed. To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements. However, the proportion of single cells expressing individual genes also carries information and should be evaluated.

@@

\textit{Flow cytometry}:
The primary ICS data set is from a  trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen with 120 individuals (98 vaccines and 20 placebo recipients, see supplementary material~\ref{supp:statpublished}). The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T--cell subsets. Here, we analyzed the vaccinees (n=98) at two time points as a pooled data set (day 0 where no response is expected, and at day 182, the primary immunological endpoint).

Simulation studies were based on hyper parameter estimates from a dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:statpublished} for details)~\cite{Peiperl:2010ej}. The statistical analysis of ICS data in the published trial is described in the original manuscript and outlined in the supplementary information (supplementary material~\ref{supp:statpublished})~\cite{Peiperl:2010ej}. The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of flow-sorted CD8--positivie T-cells from sixteen individuals. T-cells isolated by flow cytometry from sixteen individuals were stimulated in blocks of four individuals with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post--stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from the simulated samples  was compared to paired, unstimulated controls.~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al. could you fill this in?}.

@@

\subsection{Beta-Binomial Model for One Marker}
\label{sec:DE}
For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,

@@

\subsection{Parameter estimation}
\label{sec:estimation}
Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, to simplify our calculations. Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (supplementary material), and are given by,
 \begin{align}
  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
	&=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
 \end{align} 
and
\begin{align}
	\begin{split}
\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}.\label{model2:unconstrained}
\end{split}
\end{align}
Assuming that the missing data, the $z_i$'s, are known, we define the complete log-likelihood as follows,
\begin{equation}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w)
\label{eq:cll}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated.

@@

\subsubsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
Details about the algorithms are given in supplementary material. We used the method of Raftery and Lewis \citep{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

@@

The MMOSA model had higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non--responders when the magnitude of response was small (i.e. ENV-1-PTEG stimulation) (Figure~\ref{fig:HVTN065} A),  (AUC=0.954 for MIMOSA vs. 0.923 for Fisher's exact test vs. 0.827 for LRT vs. 0.775 for ranked fold change). When the magnitude of response was larger (GAG-1-PTEG stimulation), MIMOSA performed as well as Fisher's exact test, and better than other competing methods (Figure~\ref{fig:HVTN065} C) (AUC=0.982 for MIMOSA vs. 0.978 for Fisher's exact test vs. 0.877 for LRT vs. 0.784 for ranked fold change). Irrespective of the magnitude of the response to stimulation, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B,D). These results are consistent for other cytokines (see supplementary material~\ref{supp:HVTN065Results}, supplementary figures~\ref{fig:HVTN065Results} and~\ref{fig:HVTN065ResultsCont}).

\subsection{Fluidigm}
\subsubsection{MIMOSA Identifies Stimulation-Specific Patterns of Gene Expression in Fluidigm Single--Cell Data}
We applied the MIMOSA model to Fluidigm single--cell gene expression data from CD8+ T--cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples. The unconstrained MIMOSA model was fit to each stimulation. From the model we calculated the posterior probabilities of response for each gene, the posterior differences in the proportion of cells expressing each gene between stimulated and unstimulated conditions, as well as the posterior log ratio of the proportion of cells expressing each gene in stimulated and unstimulated conditions. The data are presented in (Figure~\ref{fig:fluidigm} A-C). Importantly, we see that MIMOSA identifies stimulation--specific differences in the proportions of cells expressing each gene, while preserving inter--subject variability (Figure~\ref{fig:fluidigm}). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions and the posterior log fold changes of proportions(Figure~\ref{fig:fluidigm} B,C).

@@

\subsection{Parameter estimation}
Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see supplementary material). For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
\begin{align}
\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
\end{align}
where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$. Similarly the marginal likelihood for the alternative model is given by 
\begin{align}
\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}.\label{eq:postmult}
\end{align}
The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger. In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than 2. The EM algorithms becomes very dependent on the initial values, and can even fail to converge when good initial values are provided. Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and can be used even when $K$ is large. More details about both algorithms are given in supplementary material.

@@

Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we performed simulations (five--dimensional data) to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, Fisher's exact test performed marginally on each dimension of the data and p--values combined through Fisher's method ($\chi^2=-2\sum_{i=1}^k\log(p_i)$), as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C). These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data (Figure~\ref{fig:mvsimulations} B).

@@

Importantly, the analysis of real--world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}). Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination, the model is naive to vaccine time point. Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre--vaccination) and 182 (post--vaccination) than the current standard approach (Fisher's exact test), or non--Bayesian alternatives such as the likelihood ratio test or ranking by log--fold change (Figure~\ref{fig:HVTN065} A,C). Our approach treats all day 0 observations as true negatives and all day 182 observations as true positives, yet we know that not all vaccine recipients are likely to exhibit a vaccine response on day 182. None the less, this shortcoming, at worst, leads us to underestimate the true sensitivity of MIMOSA, and does not negatively impact the comparison. Additionally, MIMOSA was found to be a better fit to real--world ICS data from vaccine trials than other analysis approaches, as evidenced by more accurate estimates of true false discovery rate (Figure~\ref{fig:HVTN065} B,D).

Although ICS is the motivating example for the MIMOSA model, it can be applied any type of single--cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions. To date, the typical approach to Fluidigm single cell gene expression analysis has been focused on identifying differences in the continuous part of the signal, corresponding to the $C_t$ values and the ``missing'' data from cells that don't express a given gene has generally been ignored, or used as part of ad--hoc pre-filtering\cite{Flatz:2011jb}. The ability of MIMOSA to identify stimulation--specific expression patterns in single--cell gene expression data demonstrates not only the broader utility of the method, but importantly, also demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm} A-C). MIMOSA explicitly models and performs inference on the observed positive cell counts between different conditions. Additionally, extending the model to take into account multiple  conditions is straightforward.

Detecting differences in polyfunctional cell populations (i.e. identifying changes in cell populations that co--express multiple proteins, cytokines, or genes), is important in vaccine research, since it allows the identification of more precisely defined, more homogeneous cell popualtions~\cite{Milush:2009bz}. These polyfunctional cell populations may be correlated with differential vaccine response or efficacy and have the potential to inform vaccine development and trial design. What's more, changes in these polyfunctional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C). Existing methods for identifying polyfunctional cytokine profiles from ICS data either separately test individual combinations of cytokines, or perform global tests between groups of individuals even when the populations under study are known to be heterogeneous (i.e. mixtures of responders and non--responders).  Sometimes these tests are entirely empirical in nature (i.e. thresholds based on fold change or some factor over background), and these methods generally do not share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}. As others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account the cytokine combinations of interest~\cite{Nason:2006dx}. Here we have shown that MIMOSA has higher sensitivity and specificity than these competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Figure~\ref{fig:mvsimulations} A,C), and the model generally provides a better fit to the single--cell assay count data arising from studies with these types of experimental designs (Figure~\ref{fig:mvsimulations} B).

\section{Conclusions}
We have developed a mixture model framework that we call MIMOSA .for identifying differences between treatment conditions in paired observations of cell counts from a variety of single--cell assays that frequently arise in the context of vaccine trials and immunological studies. Our model can be applied in a univariate or multivariate fashion, and we have shown that it has greater sensitivity and specificity than other typical approaches used to analyze the type of count--based data derived from single cell assays, including Fisher's exact test, the non-Bayesian likelihood ratio test, or ad--hoc rank based approaches. The software is implemented in R and C++, and is freely available from GitHub (\url{http://www.github.org/finak/MIMOSA}).

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
%   \includegraphics[width=.75\columnwidth]{Figures/12} 
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.6\columnwidth]{Figures/2}};
    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.6\columnwidth]{Figures/12}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
    \node at (0,-0.25) [font=\tiny\sffamily] {C} ;
    \node at (4.6,-0.25) [font=\tiny\sffamily] {D} ;
\end{tikzpicture}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) GAG-1-PTEG and expressing IL2 and IFNg. ROC and FDR plots of other cytokine combinations can be found in the supplementary material.}
\label{fig:HVTN065}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
\node[anchor=south west, inner sep=0] at (4.6,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
\node[anchor=south west, inner sep=0] at (9.1,0) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot3.pdf}};
\node at (0,3) [font=\tiny\sffamily] {A} ;
\node at (4.5,3) [font=\tiny\sffamily] {B} ;
\node at (8.95,3) [font=\tiny\sffamily] {C} ;
\end{tikzpicture}
\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a significant decrease, green a significant increase, relative to the control). Columns and rows are clustered based on these signed posterior probabilities. B) The $log_2$ ratio of the proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. C) The difference in the proportion of cells expressing each gene in the stimulated vs. control samples. Ordering of the rows and columns is preserved as in A). The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10).}

\label{fig:fluidigm}
\end{figure}
